---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    pandoc_args: [
      "+RTS", "-K64m",
      "-RTS"
    ]
    keep_md: true
---

### Loading and preprocessing the data
```{r setup, echo = TRUE, include=FALSE, warning=FALSE, message=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(scales)
knitr::opts_chunk$set(cache=TRUE)
```

```{r}
dataset <- read.csv('activity.csv', header=TRUE, sep=",", na.strings = "NA")
head(dataset)
```

What is the total number of steps taken per day?
Remove NA values from the steps variable (column)

```{r}
rawdata <- dataset
dataset <- dataset[!is.na(dataset$steps), ]
dataset <- dataset[!is.na(dataset$date), ]
head(dataset)
```

Group the dataset dy date

```{r}
gdata <- group_by(dataset, date)
head(gdata)
```

Summarise the data by date with total number of steps per given day

```{r}
total_steps <- summarise(gdata, tsteps = sum(steps))
head(total_steps)
```

Plot the hisogram for total number of steps

```{r}
plt <- qplot(tsteps, data = total_steps, geom="histogram")
plt <- plt + labs(x = "Total Steps Per Day")
plt
```

### What is the mean total number of steps taken per day?

```{r}
mtotal <- sum(total_steps$tsteps)/nrow(total_steps)
mtotal <- format(mtotal, digits = 2)
```

Mean total number of steps taken per dat is `r mtotal`

What is the median total number of steps taken per day?

```{r}
medsteps <- median(total_steps$tsteps)
medsteps <- format(medsteps, digits = 2)
```

Median total number of steps taken per dat is `r medsteps`

## What is the average daily activity pattern?


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
Group the data set by interval and calculate the average number of steps per interval

```{r}
gidata <- group_by(dataset, interval)
sdata <- summarise(gidata, avesteps = mean(steps))
plot2 <- qplot(interval, avesteps, type="l", data = sdata)
plot2 <- plot2 + labs(x = "Interval")
plot2 <- plot2 + labs(y = "Average Number of Steps Taken")
maxInterval <- sdata[sdata$avesteps == max(sdata$avesteps),]$interval
plot2 + geom_vline(xintercept = maxInterval) + geom_hline(yintercept = max(sdata$avesteps))
```

Maximum number of average steps take at interval `r maxInterval`


```{r}
comp <- complete.cases(rawdata)
countnotcompelte <- length(comp[comp == FALSE])
```


Total number of rows that is not complete is `r countnotcompelte`

## Imputing missing values

Filling all the NA values with zero
```{r}
fillNAdata <-rawdata
fillNAdata[which(is.na(fillNAdata$steps)), ]$steps <- 0

gfillNAdata <- group_by(fillNAdata, date)
total_gfillNAdata <- summarise(gfillNAdata, tsteps = sum(steps))
plt3 <- qplot(tsteps, data = total_gfillNAdata, geom="histogram")
plt3 <- plt3 + labs(x = "Total Steps Per Day")
plt3
```

```{r}
mNAtotal <- sum(total_gfillNAdata$tsteps)/nrow(total_gfillNAdata)
mNAtotal <- format(mNAtotal, digits = 2)

```
Mean total number of steps taken per dat is `r mNAtotal`

```{r}
meNAdsteps <- median(total_gfillNAdata$tsteps)
meNAdsteps <- format(meNAdsteps, digits = 2)
```

Median total number of steps taken per dat is `r meNAdsteps`

By altering the NA values in the data set, the median and mean value has changed

## Are there differences in activity patterns between weekdays and weekends?

Are there differences in activity patterns between weekdays and weekends?

```{r}
fillNAdata$date <- as.Date(fillNAdata$date)
weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
fillNAdata$day <- factor((weekdays(fillNAdata$date) %in% weekdays1), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
weekendData <- fillNAdata[fillNAdata$day == "weekend", ]
weekdataData <- fillNAdata[fillNAdata$day == "weekday", ]
gweekendData <- group_by(weekendData, interval)
sgweekendData <- summarise(gweekendData, avesteps = mean(steps))
gweekdataData <- group_by(weekdataData, interval)
sgweekdataData <- summarise(gweekdataData, avesteps = mean(steps))
```

Weekend Data
```{r}
plt4 <- qplot(interval, avesteps, type="l", data = sgweekendData, geom="line")
plt4 <- plt4 + labs(x = "Weekend Intervals") + labs(y = "Average Seps")
plt4
```

Weekday Data
```{r}
plt5 <- qplot(interval, avesteps, type="l", data = sgweekdataData, geom="line")
plt5 <- plt5 + labs(x = "Weekday Intervals") + labs(y = "Average Steps")
plt5
```
